# Reproducible Research : Course Prokect 1
by Suvrorup Mukherjee
27th May, 2021

## Introduction
This is my submission for the Course Project 1 for the May 2021 version of the MOOC ["Reproducible Research" - John Hopkins Univeristy]

## Required R packages
The following R packages need to be loaded for the course work. 
```{r requiredPackages, result = FALSE}
require(data.table)
require(ggplot2)
require(plyr)
```

## Loading and Preprocessing the data

### 1. Load the data
```{r loadData}
activity <- read.csv("activity.csv")
head(activity)
```

### 2. Preprocess the data 
The dates are stored as characters. We can verify this by :
```{r preprocess1}
class(activity$date)
```

Hence we convert it into Date type.
```{r preprocess2}
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?

### Creating new table for total steps taken every day
We can see from the table the total steps taken everyday
```{r dailySteps_data}
dailySteps_data <- ddply(activity, .(date), summarise, steps = sum(steps, na.rm = TRUE))
head(dailySteps_data)
```

### Histogram for the steps taken each day 
```{r dailySteps_plot}
hist(dailySteps_data$steps, xlab = "Total steps taken each day", main = "Histogram of total steps")
```

## Mean and median of total steps taken per day
```{r mean_dailysteps}
mean(dailySteps_data$steps)
```

```{r median_dailysteps}
median(dailySteps_data$steps)
```


## What is the average daily activity pattern?

### Create a new data table for mean steps taken at each time of the day 

```{r intervalSteps_data}
intervalSteps_data <- ddply(data, .(interval), summarise, steps = mean(steps, na.rm = TRUE))
head(intervalSteps_data)
```

### Time series plot of average steps per day against intervals

```{r average_plot}
plot(intervalSteps_data$interval, intervalSteps_data$steps, type = "l")
```

### Which interval contains the maximum number of steps?

```{r intervalsteps_max}
intervalSteps_data[which.max(intervalSteps_data$steps), "interval"]
```


## Imputing missing values 

## The number of rows with missing values in the dataset
```{r num_na}
sum(!complete.cases(activity))
```

### Devise a strategy to fill in all the missing values
I'll fill the missing values with the rounded mean of the 5-minute interval.

### New dataset with missing values filled in
```{r filled_missingValues}
mv <- activity
for(i in seq(dim(mv)[1])){
  if(is.na(mv[i,"steps"])){
    mv[i, "steps"] = intervalSteps_data[which(intervalSteps_data$interval == mv[i, "interval"]), "steps"]
  }
}
head(mv)
```

### Histogram of total number of steps taken each day
```{r hist_natotal}
replacedvalue_steps <- ddply(mv, .(date), summarise, steps = sum(steps, na.rm = TRUE))
hist(replacedvalue_steps$steps, xlab = "Total steps taken each day", main = "Histogram of total steps in the new set")
```

### New Mean and Median 
We notice that the mean has increased, while the median is almost the same.
The mean and median are equal if this strategy is followed.
```{r mean_newdailysteps}
mean(replacedvalue_steps$steps)
```

```{r median_newdailysteps}
median(replacedvalue_steps$steps)
```

## Are there differences between weekdays and weekends?

### Create new factor variable in the datasteps 
```{r day_factorVariable}
mv$wday <- ifelse(wday(mv$date)<6, "weekday", "weekend")
mv$wday <- as.factor(mv$wday)
head(mv)
```

### Multipanel plot for the mean steps taken for each interval
```{r wday_plot}
wday_intervalSteps <- ddply(mv, .(interval, wday), summarise, steps = mean(steps, na.rm = TRUE))
ggplot(wday_intervalSteps, aes(x = interval, y = steps)) + geom_line() + xlab("time") + ggtitle("Number of Steps over an Average Day") + facet_grid(wday~.)
```
